\\Large Taking uncertainty seriously  \\newline \\normalsize A  Bayesian

### 2021-08-03 paperDraft/paperDraft.Rmd
r

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
print(list(factor = factor,numerator = numerator, bias = numerator / factor))
```
\normalsize

>>>>

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
whole <- rbind(t1,t2,t3,t4)
rawsd <- sd(c(whole$A,whole$B))
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
rawsd/factor
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
print(list(factor = factor,numerator = numerator, bias = numerator / factor))
```
\normalsize

### 2021-08-03 presentationESSLLI21/presentationESSLLI.Rmd
Adds small fixes
r

---
title: "\\large Taking uncertainty in word embedding bias estimation seriously \\- a  Bayesian
  approach"
author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
date:  "ExpSem2021, ESSLLI"
fontsize: 10pt
output:  
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
classoption: x11names, dvipsnames, bibspacing,natbib
urlcolor: blue
bibliography: [../references/cosineReferences1.bib]
csl: ../references/apa-6th-edition.csl 
link-citations: yes
---

@@

## Cosine-based measures of bias - visual example

### Stereotypical lists

@@

- masculine occupations:  "maestro",  "captain", "architect", etc.

@@

@@

### Example: direct bias

- The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$

<!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->

-  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is:
<!-- ($c$ is a parameter determining how strict we want to be): -->

\vspace{-2mm}

\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

@@

[@Bolukbasi2016Man]

@@

-  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X, $Y$ are protected group words

@@

@@

@@

We investigate the uncertainty that arises from raw sample sizes

@@

We make the subclasses clear, add human neutral predicates and neutral predicates for control

@@

- It throws away information about sample sizes
- It removes variation which may result in false confidence

##  Cosine distance and word connection visualization

### Analysis of a word "muslim"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)
#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")+theme_tufte(base_size = 20)
```

## Cosine distance and word connection visualization

### Analysis of a word "priest"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")+theme_tufte(base_size = 20)
```

## Advantages of including uncertainty

- It enables one to directly observe the influence of sample sizes
- It may influence risk assessment and decision making
- ...

@@

p <- ggplot(data = data.frame(distance = dis), 
            mapping = aes(x = distance))+theme_tufte()

@@

##  Dataset-level HPDIs coefficients

### Religion coefficients

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.55\textwidth}
  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
   \end{minipage}
  \end{figure}

@@

##  Dataset-level HPDIs coefficients

### Gender coefficients

\begin{center}
\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
\end{minipage}

\end{figure}

@@

## Uncertainty included in bias detection

### Closer look at Gender class

\includegraphics[width=8cm]{../images/visGenderReddit.png}

## Summary of the research

-

@@

- including contrasts in Bayesian model
- downstream tasks
- applying uncertainty to WEAT metric
- testing AIT dataset

>>>>

---
title: "\\Large Taking uncertainty seriously  \\newline \\normalsize A  Bayesian
  approach to  word embedding bias estimation "
author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
date:  "ExpSem2021, ESSLLI"
fontsize: 10pt
output:  
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
classoption: x11names, dvipsnames, bibspacing,natbib
urlcolor: blue
bibliography: [../references/cosineReferences1.bib]
csl: ../references/apa-6th-edition.csl 
link-citations: yes
---

@@

### A visual example

@@

- masculine occupations:  "maestro",  "captain", "architect", "boss", etc.

@@

<!-- ## Cosine-based measures of bias -->

<!-- ### Example: direct bias -->

<!-- - The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ -->

<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  --> -->

<!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
<!-- <!-- ($c$ is a parameter determining how strict we want to be): --> -->

<!-- \vspace{-2mm} -->

<!-- \begin{align} -->
<!-- \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert } -->
<!-- \end{align} -->

<!-- \footnotesize  -->

<!-- [@Bolukbasi2016Man] -->

@@

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%"}
words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
weat <- data.frame(words,x,y)

ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.018, size = 7 )+
    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2), , linetype =2, arrow=arrow(), size=.05, color="blue") +
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
             label = c(".7", ".6","1.1", "1.2") , color="black", 
             size=7,  fontface="bold")+
    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
           size=10,  fontface="bold")+xlim(0,1.3)+
  theme_void(base_size = 25)
```
\normalsize

@@

- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{1.2+1.1}{2} = 0.65-1.15= -0.5$

- $s_2 = s(man,A,B) = -.3$, \linebreak  $s_3 = s(woman,A,B) = .2, s_4 = s(she, A, B) = .3$

\normalsize 
\begin{align*}
\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx - 1.95
\end{align*}

@@

-  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X$, $Y$ are protected group words

@@

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "50%"}
words <- c("jew", "muslim", "christian", "terrorist","dirty","conservative","familiar","greedy","cheap")
x <- c(.1,.1,.1,.4,.4,.7,.7,1.1,1.1)
y <- c(.4,.3,.2,.1,.2,.3,.4,.1,.2)
mac <- data.frame(words,x)

ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.02, size = 9 )+
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1), , linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), , linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)

```
\normalsize

\footnotesize 
\begin{align*}
 s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
& \vdots \end{align*}

\vspace{-6mm}

\normalsize 
\begin{align*}
 MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
\end{align*}

@@

- Sensitive to the choice of protected words, capitalization, distance measures and embedding methods [@zhang2020robustness]

@@

We statistically gauge the uncertainty that arises from raw sample sizes

@@

We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).

@@

### Key conceptual issues

- It throws away information about sample sizes
- It ignores variation in the raw data, which leads to false confidence

### Our simulation

Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(0,.05)$, you still can get really high WEAT!

### One simulation

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
set.seed(123)
t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
xlimits <- c(-.3,.3)

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
```
\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
whole <- rbind(t1,t2,t3,t4)
rawsd <- sd(c(whole$A,whole$B))
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
```
\normalsize
\pause

- Raw sd in data is `r round(rawsd,3)`
- The sd of means is `r round(factor,3)`
- The WEAT score is `r round(numerator/factor,3)`
- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!

### 50k simulations (same parameters)

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
set.seed(123)
biasesNull <- numeric(50000)
for(i in 1:50000){
t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))

factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
biasesNull[i]  <- numerator / factor
}

ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte(base_size = 15)+labs(title="50k biases for identical means and sd =.05")+ xlab("bias")
```
\normalsize

##   Cosine distance and  connection type

### Distances for  "muslim"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)
#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
```

## Cosine distance and  connection type

### Distances for  "priest"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
```

## Advantages of the Bayesian way

- Direct impact of sample sizes
- Straightforward interpretation in terms of posterior probabilities
- Freedom to choose granularity level
- More honest risk assessment and decision making

@@

p <- ggplot(data = data.frame(distance = dis), 
            mapping = aes(x = distance))+theme_tufte(base_size = 15)

@@

## Bayesian model architecture

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(rethinking)
options(buildtools.check = function(action) TRUE )
religionCoefs <- ulam(
  alist(
    cosineDistance ~ dnorm(mu,sigma),
    mu <- m + co[con],
    m ~ dnorm(1,.5),
    co[con] ~ dnorm(0,.5),
    sigma ~ dcauchy(0,1)
  ),
  data = religion,
  chains=2 , iter=8000 , warmup=1000, 
  log_lik = TRUE
)
```
\normalsize

##   Dataset-level coefficients

### Religion with 89\%-compatibility intervals (HPDI)

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.55\textwidth}
  \includegraphics[width=6cm]{../images/religionCoeffs.jpeg}
   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
   \begin{itemize}
   \item All HPDIs overlap with 0 
   \item Differences between classes are relatively small
   \item Coefficients for Race are similar
   \end{itemize}
   \end{minipage}
  \end{figure}

@@

##  Dataset-level  coefficients

### Gender with 89\%-compatibility intervals (HPDI)

\begin{center}
\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=6cm]{../images/genderCoeffs.jpeg}
\end{minipage}
\begin{minipage}{0.4\textwidth}\footnotesize
   \begin{itemize}
   \item Associated and different are away from 0
   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
   \item Differences between classes are still relatively small
   \end{itemize}
   \end{minipage}
\end{figure}

@@

## Bayesian model architecture

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(rethinking)
options(buildtools.check = function(action) TRUE )
religionCoefs <- ulam(
  alist(
    cosineDistance ~ dnorm(mu,sigma),
    mu <- m[pw] + co[con],
    m[pw] ~ dnorm(1,.5),
    co[con] ~ dnorm(0,.5),
    sigma ~ dcauchy(0,1)
  ),
  data = religion,
  chains=2 , iter=8000 , warmup=1000, 
  log_lik = TRUE
)
```
\normalsize

## Word-level coefficients

\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7.5cm]{../images/visReligionReddit.png}
\end{minipage}
\begin{minipage}{0.4\textwidth}\footnotesize

\vspace{-4cm}

   \begin{itemize}
   \item Most intervals overlap with control groups
   \item Often not too much diffence between associated and different
   \end{itemize}
   \end{minipage}
\end{figure}

## Word-level coefficients

\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7.5cm]{../images/visGenderReddit.png}
\end{minipage}
\begin{minipage}{0.4\textwidth}\footnotesize

\vspace{-4cm}

   \begin{itemize}
   \item Male attributes strong co-occurrence with female attributes
   \item Sometimes different is stronger than associated
   \item Almost no overlap with control groups
   \end{itemize}
   \end{minipage}
\end{figure}

## Word-level coefficients

\begin{figure}[!htb]\centering
  \begin{minipage}{0.55\textwidth}
\includegraphics[width=7.5cm]{../images/visRaceReddit.png}
\end{minipage}
\begin{minipage}{0.4\textwidth}\footnotesize

\vspace{-4cm}

   \begin{itemize}
   \item A lot of variation between races

   \item Often not much difference between associated and different
   \end{itemize}
   \end{minipage}
\end{figure}

@@

- Including contrasts in Bayesian calculation
- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots?)
- Downstream tasks and connection with intrinsic evaluation
- Testing data from the original Implicit Association Test (IAT) 
- Applying uncertainty to WEAT and better word lists
- Looking at other similarity measures
